One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-06-02 papers/1_dmatpol/paper-r1.tex
Table added, revised equations, added discussion

\title{Density Matrix Susceptibility Tensors} %Title of paper

@@

${\bf r}_i = \tbraket{i}{\hat{\bf r}}{i}$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parametereized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w} = \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i} + C_{\beta i} {\bf b}_{\alpha i} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}
%
\Big\}  \;.
\end{multline}
%
Hence, the 3D vector ${\bf B}^{(i)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i)}$
are unique properties of electronically unperturbed molecule.

@@

Let  
$\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatially non\hyp{}uniform distributions of the electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolizes the exact estimate).
%Let us assume that we a $\delta \overline{\bf D}^{(N)}$ 
%is the $N$th known (exact) difference one\hyp{}particle density matrix
%that was computed for the electric field distribution ${\bf F}^{(N)}({\bf r})$ chosen randomly.
Now, for each pair of the AO indices we construct the following 
parameterization of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} S^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{iw} \right) F_{iu}^{(N)} F_{iw}^{(N)} \\
                +             \sum_{uw}^{x,y,z} \left( S^{[3]}_{iu} + S^{[3]}_{iw}\right) F_{iu,iw}^{(N)}
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
The multiple parameter blocks 
$S^{[1]}$, $S^{[2]}$ and $S^{[3]}$ that are associated with electric field and its gradients
appear in the first power, allowing for linear least\hyp{}squares regression.
Note that $B_u^{(i,(10))} = S^{[1]}_{iu}$, $B_{uw}^{(i,(20))} = S^{[2]}_{iu} + S^{[2]}_{iw}$
and $B_{uw}^{(i,(01))} = S^{[3]}_{iu} + S^{[3]}_{iw}$. The square bracket superscripts 
denote the block of the parameter space. We chose symmetric forms of ${\bf B}^{(i;20)}$
and ${\bf B}^{(i;01)}$ (antisymmetric form could be constructed as well by taking the difference
of the parameters from block 2 and 3, respectively). $F_{iu}^{(N)}$ and $F_{iu,iw}^{(N)}$ 
stand for the electric field $\left[ {\bf F}({\bf r}_i)\right]_u$ and the electric field
gradient $\left[ {\BM \nabla} \otimes {\bf F}({\bf r}_i) \right]_{uw}$, respectively.

To determine the optimim set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization,
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular 
and constant.
Therefore, the exact solution for the optimal parameters is given by a Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

The theory that is necessary to validate the density matrix susceptibility model from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to Psi4 quantum chemistry program.
A few simple molecules were selected as test systems, that were subject to 
perturbations due to uniform electric fields as well as collections of point charges
generating non\hyp{}uniform electric fields. 
%For finding a general unitary transformations,
%100 random point charge distributions were selected for which an exact density matrices
%were calculated.

In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = E_{\rm nuc}({\bf F}) +
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}} \\
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{multline}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx E_{\rm nuc}({\bf F}) + 
                     2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} }  \\
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{multline}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method. Orbital localization was performed by using the Boys method. Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.

The electric field\hyp{}induced dipole and quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{uw}{\beta} \;. 
  \end{align}
\end{subequations}
%
Quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:final-model.General}
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 }
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta {\BM\upmu}^{(N)} - \overline{\delta {\BM\upmu}^{(N)}}\rvert^2 }
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta \widetilde{\BM\Theta}^{(N)} - \overline{\delta \widetilde{\BM\Theta}^{(N)}}\rvert^2 }
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
wherease the latter two contain crucial information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the Work, the Hartree\hyp{}Fock level was assumed and the 6-311++G** basis set
was used. For all the calculations of the density matrix polarization susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalized by using the L{\"o}wdin symmetric orthogonalization scheme.

@@

@@

models of the density matrix susceptibility are shown in Table~\ref{t:results}. 

@@

\begin{table*}[]
\centering
\caption{My caption}

@@

\textbf{Set}         & \multicolumn{2}{c}{\textbf{1, 0}}          &  & \multicolumn{2}{c}{\textbf{1, 0}} & \multicolumn{2}{c}{\textbf{2, 0}} & \multicolumn{2}{c}{\textbf{1, 1}} & \multicolumn{2}{c}{\textbf{2, 1}} &  & \multicolumn{2}{c}{\textbf{Polarizability}}         \\ \hline

@@

We considered three different types of point charge environment that generate
weak (0.001 -- 0.005 a.u.), 
moderate (0.001 -- 0.002 a.u.) and strong (0.002 -- 0.004 a.u.) 
electric fields around the water molecule (see the histograms of the electric field distribution
per atomic centre in Figure~XXX).
-%
-%for a set of varying point charge environments
-%with weak, moderate and strong electric fields are shown in Table~XXX.
\begin{figure*}

>>>>

\title{One-Particle Density Matrix Polarization Susceptibility Tensors}

@@

${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parametereized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w} = \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i} + C_{\beta i} {\bf b}_{\alpha i} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}
%
\Big\}  \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i)}$
are unique properties of electronically unperturbed molecule.

@@

Let  
$\left\{ {\bf F}^{(1)}({\bf r}), {\bf F}^{(2)}({\bf r}), \ldots, {\bf F}^{(N)}({\bf r}), \ldots \right\}$ 
be a set of $N_{\rm max}$ distinct and randomly sampled 
spatially non\hyp{}uniform distributions of the electric field. It is assumed that
the exact difference one\hyp{}particle density matrices (with respect to the unperturbed state)
defined as
%
\begin{equation}
 \delta \overline{\bf D}^{(N)} \equiv \overline{\bf D}^{(N)} - \overline{\bf D}^{(0)}
\end{equation}
%
are known for each sample (overline symbolizes the exact estimate).
%Let us assume that we a $\delta \overline{\bf D}^{(N)}$ 
%is the $N$th known (exact) difference one\hyp{}particle density matrix
%that was computed for the electric field distribution ${\bf F}^{(N)}({\bf r})$ chosen randomly.
Now, for each pair of the AO indices we construct the following 
parameterization of Eq.~\eqref{e:final-model.General}:
%
\begin{multline}\label{e:final-model.General.Parameters}
 \delta D^{(N)} = \sum_{i }^M \Big\{ 
                              \sum_u^{x,y,z} S^{[1]}_{iu} F_{iu}^{(N)} \\
%               + \sum_{ij}^M \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{jw} \right) F_{iu}^{(N)} F_{jw}^{(N)}
                +             \sum_{uw}^{x,y,z} \left( S^{[2]}_{iu} + S^{[2]}_{iw} \right) F_{iu}^{(N)} F_{iw}^{(N)} \\
                +             \sum_{uw}^{x,y,z} \left( S^{[3]}_{iu} + S^{[3]}_{iw}\right) F_{iu,iw}^{(N)}
                        + \ldots \Big\}
\end{multline}
%
(the Greek subscripts were omitted here for notational simplicity).
The multiple parameter blocks 
$S^{[1]}$, $S^{[2]}$ and $S^{[3]}$ that are associated with electric field and its gradients
appear in the first power, allowing for linear least\hyp{}squares regression.
Note that $B_u^{(i;10)} = S^{[1]}_{iu}$, $B_{uw}^{(i;20)} = S^{[2]}_{iu} + S^{[2]}_{iw}$
and $B_{uw}^{(i;01)} = S^{[3]}_{iu} + S^{[3]}_{iw}$. The square bracket superscripts 
denote the block of the parameter space. We chose symmetric forms of ${\bf B}^{(i;20)}$
and ${\bf B}^{(i;01)}$ (antisymmetric form could be constructed as well by taking the difference
of the parameters from block 2 and 3, respectively). $F_{iu}^{(N)}$ and $F_{iu,iw}^{(N)}$ 
stand for the electric field $\left[ {\bf F}({\bf r}_i)\right]_u$ and the electric field
gradient $\left[ {\BM \nabla} \otimes {\bf F}({\bf r}_i) \right]_{uw}$, respectively.

To determine the optimum set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization,
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular 
and constant.
Therefore, the exact solution for the optimal parameters is given by the Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and the Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** method, 
was selected as a test system which was subject to 
electrostatic perturbations due to random collections of point charges
generating non\hyp{}uniform electric fields. 
Cartesian coordinates of perturbing charges were drawn from uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).

%For finding a general unitary transformations,
%100 random point charge distributions were selected for which an exact density matrices
%were calculated.
In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = E_{\rm nuc}({\bf F}) +
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}} \\
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{multline}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx E_{\rm nuc}({\bf F}) + 
                     2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} }  \\
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{multline}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid (see Eq.~\eqref{e:dmu-dpol}). 
%, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method. Orbital localization was performed by using the Boys method. Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.
%This is because we focus in this Work on the pure polarization process. Therefore, comparison between
%the results obtained by using Eq.~\eqref{e:dE-dpol} and Eq.~\eqref{e:dE} will reveal

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;. 
  \end{align}
\end{subequations}
%
where the subscript CM denotes the center of mass of water molecule.
When the distributed polarizabilities were used instead of the polarization\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
%where ${\bf r}_{\rm CM}$ is the center of mass of water molecule.

Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:final-model.General}
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 }
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta {\BM\upmu}^{(N)} - \overline{\delta {\BM\upmu}^{(N)}}\rvert^2 }
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta \widetilde{\BM\Theta}^{(N)} - \overline{\delta \widetilde{\BM\Theta}^{(N)}}\rvert^2 }
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
wherease the latter two contain crucial information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the Work, the Hartree\hyp{}Fock level was assumed and the 6-311++G** basis set
was used. For all the calculations of the density matrix polarization susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalized by using the L{\"o}wdin symmetric orthogonalization scheme.

@@

Three different types of point charge environment that generate
weak (0.003 -- 0.007 a.u.), 
moderate (0.01 -- 0.02 a.u.) and strong (0.02 -- 0.05 a.u.) 
electric fields around the water molecule were considered (Table~\ref{t:fields}),
because similar ranges of electric field are experienced by water molecules
in liquid phase.\cite{Reischl.Kofinger.Dellago.MolPhys.2009,Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
\begin{table}%[]
\caption{{\bf Average electric fields in statistical sets of electrostatically perturbed states
of water molecule.} Each set was composed of 100 samples differing in the configuration
of 40 charges generated from a uniform distribution.
In this Table, `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields,
respectively.}
\label{t:fields}
\begin{tabular}{ldcdcdcd}
\multirow{2}{*}{\textbf{Set}} & \multicolumn{7}{c}{\textbf{Average Electric Field (a.u.)}}      \\
                              & \multicolumn{3}{c}{\textbf{Oxygen Atom}} & \textbf{} 
                              & \multicolumn{3}{c}{\textbf{Hydrogen Atom}} \\
\cline{2-4}
\cline{6-8}
\textbf{W}                    & 0.0046     & $\pm$     & 0.0019     &           & 0.0046     & $\pm$     & 0.0020     \\
\textbf{M}                    & 0.0184     & $\pm$     & 0.0075     &           & 0.0186     & $\pm$     & 0.0078     \\
\textbf{S}                    & 0.0368     & $\pm$     & 0.0150     &           & 0.0372     & $\pm$     & 0.0156    
\end{tabular}
\end{table}

@@

models of the density matrix susceptibility, as well as the conventional distributed polarizability
approach, are shown in Table~\ref{t:results}. 

@@

\begin{table*}%[]
\caption{{\bf Performance of the density matrix susceptibility tensors of water molecule 
in weak, moderate and strong electric fields.}
In this Table, `W', `M' and `S' denote `weak', `moderate' and `strong' electric fields,
respectively. $R$-squared coefficients are shown in parentheses.
Ranks of the density matrix susceptibility models are denoted as ($R_{\rm max}, P_{\rm max}$),
where $R_{\rm max}$ and $P_{\rm max}$ are the maximum ranks of the electric field 
and the electric field gradient in the expansion
from Eq.~\eqref{e:final-model.General}, respectively.

@@

\textbf{Set}         
& \multicolumn{2}{c}{\textbf{1, 0}}          
&  
& \multicolumn{2}{c}{\textbf{1, 0}} 
& \multicolumn{2}{c}{\textbf{2, 0}} 
& \multicolumn{2}{c}{\textbf{1, 1}} 
& \multicolumn{2}{c}{\textbf{2, 1}} 
&  
& \multicolumn{2}{c}{\textbf{Polarizability}}         \\ \hline

@@

The \emph{ab initio} DMS model quantitatively reproduces the interaction energy and the dipole moment 
in weak electric field (RMSE of 0.067~kcal/mol and RMSD of 3.4~mD). In stronger electric fields
the predictions become less quantitative, with high RMSE of 4.2~kcal/mol and RMSD of 14.7~mD in strong electric fields.
Induced quadrupole moment is described by the \emph{ab initio} DMS qualitatively well with $R^2$ not exceeding
60\%. Generalized DMS model of the same rank ($R_{\rm max}$=1, $P_{\rm max}$=0) performs slightly worse
for interaction energies but better for induced multipole moments, particularly decreasing the errors of induced quadrupole
calculations almost by a factor of two in strong electric fields, and by a factor of three in weaker fields.
Extending the parameter space improves the accuracy of the generalized DMS model, on overall. 
Including DMS of type ${\bf B}^{(20)}$, i.e., that interact with the square of electric field,
leads to improvement in prediction of all quantities evaluated in all electric field strengths considered in this Work. 
Interestingly, interaction energies are somewhat insensitive to addition of the quadrupolar DMS 
(${\bf B}^{(01)}$, interacting with the electric field gradient), in contrast to induced multipole moments
which exhibit noticeable improvement of accuracy. The most accurate generalized DMS model presented here 
($R_{\rm max}$=2, $P_{\rm max}$=1, which corresponds to including ${\bf B}^{(10)}$, ${\bf B}^{(20)}$ and ${\bf B}^{(01)}$)
works quantitatively well in weak and medium electric fields with $R^2$ being systematically 97.5\% and higher,
and with RMSE of 0.8~kcal/mol in moderate fields. 
However, the errors in the interaction energy
in strong fields are 3.4~kcal/mol, which is quite high as compared to the usually acceptable accuracy 
of 1~kcal/mol in routine quantum chemistry calculations. 
Recently, it was found by Fried et~al. by performing molecular dynamics simulations with a classical 
polarizable forcefield, that the average electric field experienced by water molecule in liquid phase
in room temperature is around 0.014 a.u. ($\sim$70 MV/cm).\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
This value corresponds to our moderate electric field set of samples (denoted as `M' in Tables~\ref{t:fields}
and \ref{t:results}) for which the
accuracy of the generalized DMS models is good 
(see also error distributions shown in Figure~\ref{f:results}).
Nevertheless, water molecules can experience
stronger electric fields in certain geometries of hydrogen bonded clusters, 
even up to 0.055 a.u. according to study by Reischl et~al.\cite{Reischl.Kofinger.Dellago.MolPhys.2009} 
Therefore, in strong electric fields
the present state of generalized DMS's must be further optimized in order to reach accuracy of at least
1~kcal/mol.

It is also instructive to compare the DMS models with the conventional approach based on the distributed
dipole\hyp{}dipole
polarizabilities. In terms of accuracy in interaction energy prediction, DMS models are on average one order 
of magnitude less accurate than the distributed polarizability approach, which provides errors of 0.3~kcal/mol
or less in all cases. On the other hand, DMS models better predict the induced dipole and quadrupole 
moments even in weak fields. Certainly, taking into account the distributed quadrupole polarizabilities
would greatly reduce errors of the standard multipole approach. However, it is believed that
the DMS approach, although less (but still sufficiently) accurate in terms of interaction energy, 
has an important advantage over the multipole\hyp{}based approach, i.e., it provides detailed
tensor description of the spatial polarization of the electron density distribution.

\begin{figure*}[h]
